The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-13 talk/vmil2012/paper.tex
update figure with the correct variable indices and add the line numbers to the trace operations
improve the guards in the backend section

%___________________________________________________________________________
\section{Introduction}

@@

After optimization the resulting trace is handed over to the platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward pass to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics. Operations without side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can often be merged, further reducing even more the overhead of the guard.
Figure \ref{fig:trace-compiled} shows how an \texttt{int\_eq} operation
followed by a guard that checks the result of the operation are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

\todo{Figure needs better formatting}
\begin{figure}[ht]
  \noindent
  \centering
  \begin{minipage}{1\columnwidth}
    \begin{lstlisting}[mathescape]
$b_1$ = int_eq($i_2$, 1)
guard_false($b_1$)
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
MOVEQ r8, #1
MOVNE r8, #0
CMP r8, #0
BEQ <bailout>
    \end{lstlisting}
  \end{minipage}
  \hfill
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
BNE <bailout>
...
...
...
    \end{lstlisting}
  \end{minipage}
  \caption{Result of separated (left) and merged (right) compilation of operations and guards (top).}
  \label{fig:trace-compiled}
\end{figure}

@@

First a special data
structure called \emph{low-level resume data} is created. This data structure encodes the
information about where, i.e. which register or stack location, the IR-variables required to rebuild the state will be stored when the guard is executed.
This data
structure stores the values in a succinct manner using an encoding that uses
8 bits to store 7 bits of information, ignoring leading zeros. This encoding is efficient to create and
provides a compact representation of the needed information,
to maintain an acceptable memory profile.

Second a piece of code is generated for each guard that acts as a trampoline.
Guards are implemented as a conditional jump to this trampoline in case the
guard checked fails.
In the trampoline the pointer to the
\emph{low-level resume data} is loaded and execution jumps to a generic bailout handler, also known as compensation code,
that is used to leave the compiled trace in case of a guard failure.

@@

As in previous sections the underlying idea for the design of guards is to have
a fast on-trace profile and a potentially slow one in the bailout case where
the execution takes one of the side exits due to a guard failure. At the same
time the data stored in the backend, required to rebuild the state, should be as
compact as possible to reduce the memory overhead produced by the large number
of guards, the numbers in Figure~\ref{fig:backend_data} illustrate that the
compressed encoding currently has about 15\% to 25\% of the size of of the
generated instructions on x86.

>>>>

%___________________________________________________________________________
\todo{find a better name for \texttt{low-level resume data}}
\todo{find better names for JIT front- and backend}
\section{Introduction}

@@

After optimization the resulting trace is handed over to the platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward pass to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics. Operations without side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can often be merged, further reducing the overhead of the guard.
Figure \ref{fig:trace-compiled} shows how the \texttt{int\_eq} operation
followed by a \texttt{guard_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

\todo{Figure needs better formatting}
\begin{figure}[ht]
  \noindent
  \centering
  \begin{minipage}{1\columnwidth}
\begin{lstlisting}[mathescape, numbers=right, escapechar=|, firstnumber=18]
$b_3$ = int_eq($i_5$, 1)                    |\setcounter{lstnumber}{17}|
guard_false($b_3$)                          |\setcounter{lstnumber}{-1}|
\end{lstlisting}
  \end{minipage}
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
MOVEQ r8, #1
MOVNE r8, #0
CMP r8, #0
BEQ <bailout>
    \end{lstlisting}
  \end{minipage}
  \hfill
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
BNE <bailout>
...
...
...
    \end{lstlisting}
  \end{minipage}
  \caption{Result of separated (left) and merged (right) compilation of operations and guards (top).}
  \label{fig:trace-compiled}
\end{figure}

@@

First a special data
structure called \emph{low-level resume data} is created. This data structure encodes the
information about where, i.e. which register or stack location, the IR-variables required to rebuild the state will be stored when the guard is executed.
This data
structure stores the values in a succinct manner using an encoding that uses
8 bits to store 7 bits of information, ignoring leading zeros. This encoding is efficient to create and
provides a compact representation of the needed information in order
to maintain an acceptable memory profile.

Second for each guard a piece of code is generated that acts as a trampoline.
Guards are implemented as a conditional jump to this trampoline in case the
guard check fails.
In the trampoline the pointer to the
\emph{low-level resume data} is loaded and after storing the current execution state
(registers and stack) execution jumps to a generic bailout handler, also known
as \texttt{compensation code},
that is used to leave the compiled trace in case of a guard failure.

@@

As in previous sections the underlying idea for the design of guards is to have
a fast on-trace profile and a potentially slow one in the bailout case where
the execution has to return to the interpreter due to a guard failure. At the same
time the data stored in the backend, required to rebuild the state, should be as
compact as possible to reduce the memory overhead produced by the large number
of guards, the numbers in Figure~\ref{fig:backend_data} illustrate that the
compressed encoding currently has about 15\% to 25\% of the size of of the
generated instructions on x86.
